Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-31 doc/thesis/intro.tex
More changes... i'm so unproductive

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\[
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
\]
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will attempt to approximate it using a \defn{Monte Carlo}
approach.
TODO
}}

@@

@@

{\color{blue}\uline{The resampling step was formally integrated with
\gls{SIS} by \citeauthor{doucet2000sequential} to form the first \gls{SMC}
algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
bootstrap filter proposed, the authors use a criterion based on the \gls{ESS}
the particle population to determine when resampling is necessary. }} The
\gls{ESS} of the population of particles is defined as 
\[ 
  \ESS(w) = \frac{n}{1 + \Var(w)}, 
\]
where $n$ is the number of particles in the population.
When the \gls{ESS} drops below the threshold
{\color{blue}\uline{(conventionally $n/2$~\autocite{liu2008monte})}}, particles
are resampled according to their importance weights. This results in the
removal of low-weight particles from the population, and also equalizes all the
weights. Various resampling strategies beyond the basic sampling with
replacement have been proposed~\autocite{douc2005comparison}, but we will not
discuss those here.

@@

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $x$ and outcomes $y$. A
\defn{parametric} model defines a family of possible relationships between data
and outcomes, indexed by one or more numeric parameters $\theta$. A
\defn{statistical} model describes the relationship between data and outcomes
in terms of probabilities. Statistical models define, either explicitly or
implicitly, the probability of observing $y$ given $x$ and, if the model is
parametric, $\theta$. Note that it is entirely possible to have no data $x$,
only observed outcomes $y$. In this case, a model would describe the process by
which $y$ is generated.

@@

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and can be accomplished via the Metropolis-Hastings
algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).

>>>>

@@

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
  \label{eq:post}
\end{align}
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will approximate it using a \defn{Monte Carlo} approach.
The fundamental idea behind Monte Carlo methods is succinctly expressed by
\textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions, including as the aforementioned
posterior, that are analytically intractable and from which direct sampling is
not possible } (for a review see \autocite{robert2004monte}).
\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte} \uline{is one such algorithm.}

\uline{ \Gls{SMC} is also known as the \emph{particle filter}. Rather than
sampling points one at a time from the target distribution, \gls{SMC} considers
a population of points or ``particles'', here denoted $\set{x^{(k)}}$. The
particles are associated with weights, $\set{w(x^{(k)})}$. These weighted
particles are a representation for the target distribution. For example, the 
expected value of the distribution is approximated by}
\[
  \frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),
\]
\uline{where $n$ is the number of particles. Initially, the particles do not
represent the target distribution but rather a more tractable distribution from
which direct sampling is straightforward. The word ``sequential'' is used to
describe the iterative process of perturbation, resampling, and reweighting
applied to the particles in such a way that they converge, collectively, to a
sample from the target. }

\uline{In this work, the distribution of interest is the posterior distribution
\ref{eq:post}. The particles are particular values of the parameters $\theta$
of the contact network model being studied. If we were taking a typical
Bayesian Monte Carlo approach to this problem, the particles would end up
weighted by their posterior probability and distributed in such a way that the
weighted population was a reasonable representation of $\Pr(\theta \mid T)$.
This turns out to be not straightforward in our case, as we will address in
}\cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}

@@

{\color{blue}\uline{The resampling step was formally integrated with
\gls{SIS} by \textcite{doucet2000sequential} to form the first \gls{SMC}
algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
bootstrap filter proposed, the authors use a criterion based on the \gls{ESS}
the particle population to determine when resampling is necessary. }} The
\gls{ESS} of the population of particles is defined as 
\[ 
  \ESS(w) = \frac{n}{1 + \Var(w)}, 
\]
where $n$ is the number of particles in the population.
When the \gls{ESS} drops below the threshold
{\color{blue}\uline{(conventionally $n/2$~\autocite{liu2008monte})}}, particles
are resampled according to their importance weights. This results in the
removal of low-weight particles from the population, and also equalizes all the
weights. Various resampling strategies beyond the basic sampling with
replacement have been proposed~\autocite{douc2005comparison}, but we will not
discuss those here.

@@

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $x$ and outcomes $y$. A
\defn{parametric} model defines a family of possible relationships between data
and outcomes, parameterized by one or more numeric parameters $\theta$. A
\defn{statistical} model describes the relationship between data and outcomes
in terms of probabilities. Statistical models define, either explicitly or
implicitly, the probability of observing $y$ given $x$ and, if the model is
parametric, $\theta$. Note that it is entirely possible to have no data $x$,
only observed outcomes $y$. In this case, a model would describe the process by
which $y$ is generated.

@@

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain {\color{red}\sout{whose states are indexed by
possible model parameters. The transition probability from state $\theta_1$ to
state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
model parameters. The transition density from parameters $\theta_1$ to
$\theta_2$ is taken to be}}
\[
  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and can be accomplished via the Metropolis-Hastings
algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).

### 2016-05-31 doc/thesis/methods_results.Rnw
More changes... i'm so unproductive

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of \textcite{del2012adaptive}, the
parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters were identifiable from the shape of the transmission
tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
was varied one at a time, while holding the other parameters at fixed, known
values. Contact networks were generated according to each set of parameter
values, and transmission trees were simulated over the networks. We then
evaluated how well a classifier based on the tree kernel could differentiate
the trees simulated under distinct parameter values. If the classifier's
cross-validation accuracy was high, this could be taken as an indication that
the parameter in question was identifiable in the range of values considered. 
A caveat of this preliminary analysis is that, since all parameters but one
were held at known values, nothing could be said about the identifiability of
\emph{combinations} of parameters; this issue will be explored later by jointly
estimating all parameters with \gls{ABC}.}

@@

<<mixed_peerdriven, include=FALSE>>=
  library(coda)
  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
  d <- d[iter == max(iter)]
  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
  params <- c("alpha", "m", "N", "I")
  d <- melt(d, measure.vars=params, variable.name="parameter")
  f <- function (x) {
    dens <- density(x)
    hpd <- HPDinterval(mcmc(x))
    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
         hpd.max=hpd[,"upper"])
  }
  d <- d[,f(value), by=parameter]
  setkey(d, parameter)

  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
  p <- p[iter == max(iter)]
  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
  params <- c("alpha", "m", "N", "I")
  p <- melt(p, measure.vars=params, variable.name="parameter")
  f <- function (x) {
    dens <- density(x)
    hpd <- HPDinterval(mcmc(x))
    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
         hpd.max=hpd[,"upper"])
  }
  p <- p[,f(value), by=parameter]
  setkey(p, parameter)
  options(scipen=5)
@

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
[95\% HPD] estimates for each parameter were: 
$\alpha$, 
  \Sexpr{round(d["alpha", map], 2)} 
  [\Sexpr{round(d["alpha", hpd.min], 2)} -
   \Sexpr{round(d["alpha", hpd.max], 2)}];
$I$,
  \Sexpr{d["I", round(map)]} 
  [\Sexpr{d["I", round(hpd.min)]} -
   \Sexpr{d["I", round(hpd.max)]}];
$m$,
  \Sexpr{d["m", floor(map)]} 
  [\Sexpr{d["m", floor(hpd.min)]} -
   \Sexpr{d["m", floor(hpd.max)]}];
$N$,
  \Sexpr{d["N", round(map)]} 
  [\Sexpr{d["N", round(hpd.min)]}-
   \Sexpr{d["N", round(hpd.max)]}].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  \Sexpr{p["alpha", round(map, 2)]} 
  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
   \Sexpr{p["alpha", round(hpd.max, 2)]}];
$I$,
  \Sexpr{p["I", floor(map)]} 
  [\Sexpr{p["I", floor(hpd.min)]} -
   \Sexpr{p["I", floor(hpd.max)]}];
$m$,
  \Sexpr{p["m", floor(map)]} 
  [\Sexpr{p["m", floor(hpd.min)]} -
   \Sexpr{p["m", floor(hpd.max)]}];
$N$,
  \Sexpr{p["N", floor(map)]} 
  [\Sexpr{p["N", floor(hpd.min)]} -
   \Sexpr{p["N", floor(hpd.max)]}].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

@@

%<<mixed_peerdriven, include=FALSE>>=
%  library(coda)
%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
%  d <- d[iter == max(iter)]
%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
%  params <- c("alpha", "m", "N", "I")
%  d <- melt(d, measure.vars=params, variable.name="parameter")
%  f <- function (x) {
%    dens <- density(x)
%    hpd <- HPDinterval(mcmc(x))
%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
%         hpd.max=hpd[,"upper"])
%  d <- d[,f(value), by=parameter]
%  setkey(d, parameter)

@@

%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
%  p <- p[iter == max(iter)]
%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
%  params <- c("alpha", "m", "N", "I")
%  p <- melt(p, measure.vars=params, variable.name="parameter")
%  f <- function (x) {
%    dens <- density(x)
%    hpd <- HPDinterval(mcmc(x))
%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
%         hpd.max=hpd[,"upper"])
%  p <- p[,f(value), by=parameter]
%  setkey(p, parameter)
%  options(scipen=5)

@@

%To test the effect of model misspecification, we simulated one network where
%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
%HPD] estimates for each parameter were: 
%$\alpha$, 
%  \Sexpr{round(d["alpha", map], 2)} 
%  [\Sexpr{round(d["alpha", hpd.min], 2)} -
%   \Sexpr{round(d["alpha", hpd.max], 2)}];
%$I$,
%  \Sexpr{d["I", round(map)]} 
%  [\Sexpr{d["I", round(hpd.min)]} -
%   \Sexpr{d["I", round(hpd.max)]}];
%$m$,
%  \Sexpr{d["m", floor(map)]} 
%  [\Sexpr{d["m", floor(hpd.min)]} -
%   \Sexpr{d["m", floor(hpd.max)]}];
%$N$,
%  \Sexpr{d["N", round(map)]} 
%  [\Sexpr{d["N", round(hpd.min)]} -
%   \Sexpr{d["N", round(hpd.max)]}].
%The approximate posterior distributions for this simulation are shown in
%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
%transmission tree in a peer-driven fashion, where the probability to sample a
%node was twice as high if one of its peers had already been sampled. The
%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
%$I$ = 2000. The estimated values were
%$\alpha$, 
%  \Sexpr{p["alpha", round(map, 2)]} 
%  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
%   \Sexpr{p["alpha", round(hpd.max, 2)]}];
%$I$,
%  \Sexpr{p["I", floor(map)]} 
%  [\Sexpr{p["I", floor(hpd.min)]} -
%   \Sexpr{p["I", floor(hpd.max)]}];
%$m$,
%  \Sexpr{p["m", floor(map)]} 
%  [\Sexpr{p["m", floor(hpd.min)]} -
%   \Sexpr{p["m", floor(hpd.max)]}];
%$N$,
%  \Sexpr{p["N", floor(map)]} 
%  [\Sexpr{p["N", floor(hpd.min)]} -
%   \Sexpr{p["N", floor(hpd.max)]}].
%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
%of these results were in line with estimates obtained on other simulated
%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
%$\alpha$ was somewhat lower than typical.

%\subsubsection*{Effect of parameters on power-law exponent}
%
%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
%to the observed distribution of \gls{gamma} values. The coefficients are
%interpretable as multiplicative effects.
%
%\begin{table}
%    \centering
%    \input{\tablepath/pa-gamma-glm}
%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
%    exponent \gls{gamma} and \gls{BA} model parameters.}
%    \label{tab:glm}
%\end{table}

>>>>

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive}, \uline{the
parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters were identifiable from the shape of the transmission
tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
was varied one at a time, while holding the other parameters at fixed, known
values. Contact networks were generated according to each set of parameter
values, and transmission trees were simulated over the networks. We then
evaluated how well a classifier based on the tree kernel could differentiate
the trees simulated under distinct parameter values. If the classifier's
cross-validation accuracy was high, this could be taken as an indication that
the parameter in question was identifiable in the range of values considered. 
A caveat of this preliminary analysis is that, since all parameters but one
were held at known values, nothing could be said about the identifiability of
\emph{combinations} of parameters; this issue will be explored later by jointly
estimating all parameters with \gls{ABC}} (\cref{subsec:abc}).

@@

<<mixed_peerdriven, include=FALSE>>=
  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
  d <- d[iter == max(iter)]
  #d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
  params <- c("alpha", "m", "N", "I")
  d <- melt(d, measure.vars=params, variable.name="parameter")
  f <- function (x, wt) {
    hpd <- wtd.hpd(x, wt/sum(wt))
    list(pmean=wtd.mean(x, wt, normwt=TRUE), 
         hpd.min=hpd[1], 
         hpd.max=hpd[2])
  }
  d <- d[,f(value, weight), by=parameter]
  setkey(d, parameter)

  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
  p <- p[iter == max(iter)]
  #p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
  params <- c("alpha", "m", "N", "I")
  p <- melt(p, measure.vars=params, variable.name="parameter")
  p <- p[,f(value, weight), by=parameter]
  setkey(p, parameter)
  options(scipen=5)
@

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
[95\% HPD] estimates for each parameter were: 
$\alpha$, 
  \Sexpr{round(d["alpha", pmean], 2)} 
  [\Sexpr{round(d["alpha", hpd.min], 2)} -
   \Sexpr{round(d["alpha", hpd.max], 2)}];
$I$,
  \Sexpr{d["I", round(pmean)]} 
  [\Sexpr{d["I", round(hpd.min)]} -
   \Sexpr{d["I", round(hpd.max)]}];
$m$,
  \Sexpr{d["m", floor(pmean)]} 
  [\Sexpr{d["m", floor(hpd.min)]} -
   \Sexpr{d["m", floor(hpd.max)]}];
$N$,
  \Sexpr{d["N", round(pmean)]} 
  [\Sexpr{d["N", round(hpd.min)]}-
   \Sexpr{d["N", round(hpd.max)]}].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  \Sexpr{p["alpha", round(pmean, 2)]} 
  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
   \Sexpr{p["alpha", round(hpd.max, 2)]}];
$I$,
  \Sexpr{p["I", floor(pmean)]} 
  [\Sexpr{p["I", floor(hpd.min)]} -
   \Sexpr{p["I", floor(hpd.max)]}];
$m$,
  \Sexpr{p["m", floor(pmean)]} 
  [\Sexpr{p["m", floor(hpd.min)]} -
   \Sexpr{p["m", floor(hpd.max)]}];
$N$,
  \Sexpr{p["N", floor(pmean)]} 
  [\Sexpr{p["N", floor(hpd.min)]} -
   \Sexpr{p["N", floor(hpd.max)]}].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

@@

%\subsubsection*{Effect of parameters on power-law exponent}

@@

%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
%to the observed distribution of \gls{gamma} values. The coefficients are
%interpretable as multiplicative effects.

@@

%    \input{\tablepath/pa-gamma-glm}
%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
%    exponent \gls{gamma} and \gls{BA} model parameters.}
%    \label{tab:glm}
